---
title: "Microbial Mapping - Data Import and Clean"
output: html_notebook
---

# Document Summary
This code is for importing and cleaning outputs from the HPC metaflowmics pipeline.
Pipeline output files are read in, written out as cleaned flat files, and read back in to create a phyloseq object.


```{r}
library(data.table)
library(phyloseq)
if(getwd() != "/home/sean/Documents/Bioinformatics/Projects/Waimea/microbial_mapping/main")
  {
setwd("/home/sean/Documents/Bioinformatics/Projects/Waimea/microbial_mapping/main")
}

```



## Read In and Join Metadata
Use data.table package to read in and clean metadata.
pipeline output data files for 16s and its


```{r}


# sample metadata
raw_samples  <- fread("../data/raw/sample/unique_waimea_hiseq_samples_emp.tsv",
                      sep = "\t", header = T)

# library prep metadata
prep_16s <- fread("../data/raw/sample/combined_hiseqs_bioblitz_mapping - 16S_prep_info.tsv",
                  sep = "\t", header = T)

prep_its <- fread("../data/raw/sample/combined_hiseqs_bioblitz_mapping - ITS_prep_info.tsv",
                  sep = "\t", header = T)

# joined prep and sample metadata
meta_16s <- merge(prep_16s, raw_samples, all.x=T)
meta_its <- merge(prep_its, raw_samples, all.x=T)

meta_all <-rbindlist(list(meta_16s,meta_its))

fwrite(meta_16s,"../data/processed/interim/combined_prep_sample_metadata_16s.csv")
fwrite(meta_its,"../data/processed/interim/combined_prep_sample_metadata_its.csv")
fwrite(meta_all,"../data/processed/interim/combined_prep_sample_metadata_16sandits.csv")

# 16s data
abund_16s_file <- "../data/raw/abundance_table_97.shared"
tax_16s_file  <- "../data/raw/annotations_97.taxonomy"
meta_16s_file <- "../data/processed/interim/combined_prep_sample_metadata_16s.csv"


# its data

```



## Clean data
Source function for creating phyloseq objects from standard hpc output.
The function will read in files, clean them, organize them as a list and write out each component (abundance, taxonony, metadata) as a csv file. 

```{r}
source("../src/clean_and_query_16S.R")


## clean_16s_tables outputs a list of cleaned versions of each dataframe
## 'cleaning' means renaming some columns and generally tidying the pipeline outputs
## sequences are not affected
## names are checked to make sure they agree between all files
## the function writes out each cleaned file as a csv

mm_16s <- clean_16S_tables(
                  abundance_file = abund_16s_file,
                  taxonomy_file = tax_16s_file,
                  metadata_file = meta_16s_file,
                  description = "mm_16s_hiseqs",
                  output_dir = "../data/processed/cleaned",
                  id_column = "sequencing_id",
                  cull = list(min.num = 4, min.abund = 500, min.single.abund = 1000)
                  )

# write out the list of dataframes as r object
saveRDS(mm_16s, file = "../data/processed/cleaned/mm_16s.rds")

# remove variable to save memory
rm(mm_16s)
gc()

```

## Create Phyloseq Object
The function takes the csv files written out above. Each file is read in, re-organized for phyloseq object, and finally a phyloseq object is created.

```{r}
# make phyloseq
source("../src/make_phyloseq.R")

MM_16S_phy <- make_phyloseq(abund_file  = "../data/processed/cleaned/mm_16s_hiseqs_abundance_table.csv",
                            tax_file    = "../data/processed/cleaned/mm_16s_hiseqs_taxonomy_table.csv",
                            meta_file   = "../data/processed/cleaned/mm_16s_hiseqs_metadata_table.csv",
                            id_column = "sequencing_id")

ntaxa(MM_16S_phy)


# filter taxa
MM_16S_phy <- subset_taxa(MM_16S_phy, (order!="Chloroplast") &
                                         (kingdom != "Archaea") &
                                         (family != "Mitochondria") &
                                         (kingdom != "Unknown"))
ntaxa(MM_16S_phy)

nsamples(MM_16S_phy)


# filter out empty samples
MM_16S_phy <- prune_samples(sample_sums(MM_16S_phy) != 0, MM_16S_phy)

nsamples(MM_16S_phy)

# filter by project
MM_16S_phy <- subset_samples( MM_16S_phy, project_name %in% c("Keck","JoshMS"))


# write out phyloseq
saveRDS(MM_16S_phy, "../data/processed/cleaned/MM_16S_phy.rds")

# remove phyloseq object
rm(MM_16S_phy)


```


