---
title: "Microbial Mapping - NMDS"
output: html_notebook
---
# Document Summary
This code is for generating figures related to the actual sequencing data from the waimea hiseq data sets.
first and foremost, the goal is to generate a variety of nmds plots to visualize braod trends in the data.


## Set Up Environment
Set a standardized color palettes for plotting and load libraries

```{r}
library(dplyr)
library(ggplot2)
library(data.table)

# check working directory
setwd("/home/sean/documents/bioinformatics/projects/waimea/microbial_mapping/main")

# set ggplot theme
theme_set(theme_bw())

# set color palettes 
pal_d = "spectral"
scale_colour_discrete <-  function(palname=pal_d, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal_d, ...){
    scale_fill_brewer(palette=palname, ...)
}

pal_c = "c"
uniform_gradient_color <- function(palname = pal_c){
  scale_color_viridis_c(option = palname)
}

# can change these settings with viridis package
# with colour brewer, need add a line to each plot for continuous colors
# "scale_color_gradient2(low = "green", mid = "blue", high = "red")"

#  scale_colour_continuous <-  function(palname=pal_c, ...){
#    scale_colour_brewer(palette=palname, ...)
#  }
# # 
#  scale_fill_continuous <-  function(palname=pal_c, ...){
#    scale_fill_brewer(palette=palname, ...)
#  }



```






## all sample nmds
run nmds on all samples

```{r}

```

### by trophic
run nmds colored by trophic level for all samples, and split by habitat
```{r}

```

### by habitat
```{r}

```

### by host
```{r}

```

## subset nmds
```{r}

```
